### 2024-01-24 outputs/paper/paper.qmd
somehow, still awake. I HAVE A CLEAN DATASET!!!

---
title: "Exploring TTC Delays"
subtitle: "A Comprehensive Analysis"
author: Luca Carnegie
thanks: "Code and data are available at: https://github.com/lcarnegie/OpenDataToronto. Thank you to coffee, Rajan Maghera, Hannah Yu, and Sehar Bajwa for your love and support - could not have done it without you. No thanks go to Rohan for forcing me to teach myself R. :)"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---
# Introduction

# Data

# References

### 2024-01-24 scripts/02-data_cleaning.R

@@ -1,44 +1,107 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Prerequisites: 01-download_data.R
+# Purpose: Cleans the raw TTC data downloaded from OpenDataToronto
+# Author: Luca Carnegie 
+# Date: January 24, 2024
+# Contact: luca.carnegie@mail.utoronto.ca
+
+
 
 #### Workspace setup ####
 library(tidyverse)
+library(janitor)
+library(lubridate)
+
 
 #### Clean data ####
-raw_data <- read_csv("inputs/data/plane_data.csv")
+raw_data_bus <- read_csv("inputs/data/bus_data.csv")
+raw_data_streetcar <- read_csv("inputs/data/streetcar_data.csv")
+raw_data_subway <- read_csv("inputs/data/subway_data.csv")
 
-cleaned_data <-
-  raw_data |>
-  janitor::clean_names() |>
-  select(wing_width_mm, wing_length_mm, flying_time_sec_first_timer) |>
-  filter(wing_width_mm != "caw") |>
-  mutate(
-    flying_time_sec_first_timer = if_else(flying_time_sec_first_timer == "1,35",
-                                   "1.35",
-                                   flying_time_sec_first_timer)
-  ) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "490",
-                                 "49",
-                                 wing_width_mm)) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "6",
-                                 "60",
-                                 wing_width_mm)) |>
-  mutate(
-    wing_width_mm = as.numeric(wing_width_mm),
-    wing_length_mm = as.numeric(wing_length_mm),
-    flying_time_sec_first_timer = as.numeric(flying_time_sec_first_timer)
-  ) |>
-  rename(flying_time = flying_time_sec_first_timer,
-         width = wing_width_mm,
[…]

### 2024-01-24 outputs/paper/paper.qmd
introduction, done. some references, done. just need a graph or two.

@@
 title: "Exploring TTC Delays"
-subtitle: "A Comprehensive Analysis"
+subtitle: "A Multimodal Analysis"
 author: Luca Carnegie
 thanks: "Code and data are available at: https://github.com/lcarnegie/OpenDataToronto.
-Thank you to coffee, Rajan Maghera, Hannah Yu, and Sehar Bajwa for your love and support - could not have done it without you.
-No thanks go to Rohan for forcing me to teach myself R.
+Thank you to Nescafé coffee, Rajan Maghera, Egg McMuffins, Hannah Yu, and Sehar Bajwa for your love and support - could not have done it without you.
+No thanks go to Rohan in particular for not pointing out all the R he wanted us to learn that was in his textbook...
+sigh.
+I will try the appendix cocktails someday, though.
 :)"
@@
 format: pdf
+fig_caption: yes
 number-sections: true
@@
 ---
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+
+library(knitr)
+library(janitor)
+library(lubridate)
+library(opendatatoronto)
+library(tidyverse)
+library(tidyr)
+library(data.table)
+library(dplyr)
+library(kableExtra)
+```
+
 # Introduction
 
-# Data
+A key area of improvement in Canadian sustainability is in increasing the use of public transportation in big cities like Toronto.
+This makes sense since cars account for about 80% of transportation emissions [@diaetal].
+However, a major drawback preventing the widespread use of public transportation is the existence of delays within the transit system.
+These delays incentivize citizens toward using less sustainable alternatives like private cars, leading to congestion on highways and consequent increased emissions.
+The lack of timely transit services have a demonstrated impact on traffic.
+One study by Anderson et al.
+estimated a whopping 47 percent increase in highway delay when transit service ceased completely in a big city such as Los Angeles [@anderson].
+It is clear to see, then, that it is imperative that policy makers understand and fix these delays to improve sustainability.
+
+The "TTC", or Toronto Transit Commission, has been Toronto's transit system since 1921, serving the city population with subways, streetcars, buses, along with smaller offerings.
+The openly available data surrounding the delays of these various modes of transport offers opportunity to investigate the root causes of the longer-lasting delays.
+Understanding what caused a delay, when it was caused and where it was caused could give the TTC more insight into how to deliver their services to the public more efficiently and effectively, setting an examples of what an effective transit agency could look like.
+
+This paper aims to find relationships and patterns of public transit delay time across various genres of transportation.
+First, the data source and analysis employed are covered.
+Section Three then critically examines the data, delving deeper to arrive at various conclusions, as well as discussing implications and proposing new areas of exploration.
+
+# Data {#sec-data}
+
+To investigate transit delays in Toronto, data on bus [@busdata], subway [@subwaydata], and streetcar [@streetcardata] delay incidents for the year 2023 was downloaded using the OpenDataToronto R package [@odt].
+Then, the data was cleaned using R [@citeR], tidyverse [@tidyverse] and it's associated packages.
+After combining the datasets together, variables common between datasets and relevant to the analysis were selected.
+In the case of this analysis, the variables included were date (year, month, day), day of the week, Location of Delay Incident, Cause of Delay, the time delay (in minutes), Vehicle (bus, subway, streetcar), as well as the route/line (simplified to 'line') the vehicle was on when the incident happened.
+(Perhaps add seasonal analysis??)
+
+```{r}
+#| message: false
+#| echo: false
+#| label: tbl-cleaned_data
+#| tbl-cap: Sample of Cleaned Delay Data
+
+cleaned_transit_data <- read_csv(here::here("/outputs/data/cleaned_dataset.csv"))
+
+head (cleaned_homicide_data, 5) 
+```
+
+## General Delay Analysis {#subsec1-data}
+
+# Discussion {#sec-discussion}
 
 # References
+
+---
+nocite: |
+  @citeR, @ggplot, @odt, @janitor, @lubridate, @tidyverse, @dplyr, @ggplot, @testthat, @MartinWittmanLi, @diaetal, @anderson,
+---

### 2024-01-24 scripts/03-test_data.R

@@ -1,15 +1,48 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Purpose: Tests Cleaned Dataset Developed
+# Author: Luca Carnegie
+# Date: 24 January 2023
+# Contact: luca.carnegie@mail.utoronto.ca
+# Pre-requisites: 02-data_cleaning.R
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
 
 #### Test data ####
+
+dataset <- read_csv("outputs/data/cleaned_dataset.csv")
+
+# Load necessary libraries
+library(dplyr)
+library(testthat)
+
+# Assume df is your data frame
+
+# Test that the data frame has the correct columns
+test_that("Data frame has correct columns", {
+  expected_cols <- c("year", "month", "day", "weekday", "hour", "minute", "line", "vehicle", "incident", "delay")
+  expect_equal(colnames(df), expected_cols)
+})
+
+# Test that there are no missing values
+test_that("No missing values", {
+  expect_equal(sum(is.na(df)), 0)
+})
+
+# Test that numeric columns have reasonable values
+test_that("Numeric columns have reasonable values", {
+  expect_true(all(df$year >= 2000 & df$year <= 2025))
+  expect_true(all(df$month >= 1 & df$month <= 12))
+  expect_true(all(df$day >= 1 & df$day <= 31))
+  expect_true(all(df$hour >= 0 & df$hour <= 23))
+  expect_true(all(df$minute >= 0 & df$minute <= 59))
+  expect_true(all(df$delay >= 0))
+})
+
+# Test that categorical columns have expected levels
+test_that("Categorical columns have expected levels", {
+  expect_true(all(df$weekday %in% c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")))
+  expect_true(all(df$vehicle %in% c("Subway", "Bus", "Streetcar")))
+  expect_true(all(df$incident %in% c("Mechanical", "Operations", "General Delay", "Emergency")))
+})

### 2024-01-24 outputs/paper/paper.qmd
submitted something for peer review. but at least the analysis from here will be easy.

@@
 
+## General Delay Analysis {#subsec1-data}
+
 ```{r}
@@
 #| tbl-cap: Sample of Cleaned Delay Data
+#| fig-align: center
 
-cleaned_transit_data <- read_csv(here::here("/outputs/data/cleaned_dataset.csv"))
+cleaned_transit_data <- read_csv(here::here("outputs/data/cleaned_dataset.csv"))
 
-head (cleaned_homicide_data, 5) 
+sample_table <- head(cleaned_transit_data, 5)
+
+knitr::kable(sample_table)
 ```
-
-## General Delay Analysis {#subsec1-data}
 

### 2024-01-24 outputs/paper/paper.tex
submitted something for peer review. but at least the analysis from here will be easy.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Exploring TTC Delays},
  pdfauthor={Luca Carnegie},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Exploring TTC Delays\thanks{Code and data are available at:
https://github.com/lcarnegie/OpenDataToronto. Thank you to Nescafé
coffee, Rajan Maghera, Egg McMuffins, Hannah Yu, and Sehar Bajwa for
your love and support - could not have done it without you. No thanks go
to Rohan in particular for not pointing out all the R he wanted us to
learn that was in his textbook\ldots{} sigh. I will try the appendix
cocktails someday, though. :)}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{A Multimodal Analysis}
\author{Luca Carnegie}
\date{January 24, 2024}

\begin{document}
\maketitle
\begin{abstract}
First sentence. Second sentence. Third sentence. Fourth sentence.
\end{abstract}

\section{Introduction}\label{introduction}

A key area of improvement in Canadian sustainability is in increasing
the use of public transportation in big cities like Toronto. This makes
sense since cars account for about 80\% of transportation emissions (Dia
et al. 2019). However, a major drawback preventing the widespread use of
public transportation is the existence of delays within the transit
system. These delays incentivize citizens toward using less sustainable
alternatives like private cars, leading to congestion on highways and
consequent increased emissions. The lack of timely transit services have
a demonstrated impact on traffic. One study by Anderson et al.~estimated
a whopping 47 percent increase in highway delay when transit service
ceased completely in a big city such as Los Angeles (Anderson 2014). It
is clear to see, then, that it is imperative that policy makers
understand and fix these delays to improve sustainability.

The ``TTC'', or Toronto Transit Commission, has been Toronto's transit
system since 1921, serving the city population with subways, streetcars,
buses, along with smaller offerings. The openly available data
surrounding the delays of these various modes of transport offers
opportunity to investigate the root causes of the longer-lasting delays.
Understanding what caused a delay, when it was caused and where it was
caused could give the TTC more insight into how to deliver their
services to the public more efficiently and effectively, setting an
examples of what an effective transit agency could look like.

This paper aims to find relationships and patterns of public transit
delay time across various genres of transportation. First, the data
source and analysis employed are covered. Section Three then critically
examines the data, delving deeper to arrive at various conclusions, as
well as discussing implications and proposing new areas of exploration.

\section{Data}\label{sec-data}

To investigate transit delays in Toronto, data on bus (Commission
2024a), subway (Commission 2024c), and streetcar (Commission 2024b)
delay incidents for the year 2023 was downloaded using the
OpenDataToronto R package (Gelfand 2022). Then, the data was cleaned
using R (R Core Team 2022), tidyverse (Wickham et al. 2019) and it's
associated packages. After combining the datasets together, variables
common between datasets and relevant to the analysis were selected. In
the case of this analysis, the variables included were date (year,
month, day), day of the week, Location of Delay Incident, Cause of
Delay, the time delay (in minutes), Vehicle (bus, subway, streetcar), as
well as the route/line (simplified to `line') the vehicle was on when
the incident happened. (Perhaps add seasonal analysis??)

\subsection{General Delay Analysis}\label{subsec1-data}

\begin{table}

\caption{\label{tbl-cleaned\_data}Sample of Cleaned Delay Data}

\centering{

\begin{tabular}{l|l|l|l|r|l|l|r}
\hline
Date & Time & Day & vehicle & Line & Location & Incident & Min Delay\\
\hline
2023-01-01 & 02:30:00 & Sunday & Bus & 91 & WOODBINE AND MORTIMER & Diversion & 81\\
\hline
2023-01-01 & 02:34:00 & Sunday & Bus & 69 & WARDEN STATION & Security & 22\\
\hline
2023-01-01 & 03:06:00 & Sunday & Bus & 35 & JANE STATION & Cleaning & 30\\
\hline
2023-01-01 & 03:14:00 & Sunday & Bus & 900 & KIPLING STATION & Security & 17\\
\hline
2023-01-01 & 03:43:00 & Sunday & Bus & 85 & MEADOWALE LOOP & Security & 1\\
\hline
\end{tabular}

}

\end{table}%

\section{Discussion}\label{sec-discussion}

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-anderson}
Anderson, Michael L. 2014. {``Subways, Strikes, and Slowdowns: The
Impacts of Public Transit on Traffic Congestion.''} \emph{American
Economic Review} 104 (9): 2763--96.
\url{https://doi.org/10.1257/aer.104.9.2763}.

\bibitem[\citeproctext]{ref-busdata}
Commission, Toronto Transit. 2024a. {``TTC Bus Delay Data.''}
\url{https://open.toronto.ca/dataset/ttc-bus-delay-data/}.

\bibitem[\citeproctext]{ref-streetcardata}
---------. 2024b. {``TTC Streetcar Delay Data.''}
\url{https://open.toronto.ca/dataset/ttc-streetcar-delay-data/}.

\bibitem[\citeproctext]{ref-subwaydata}
---------. 2024c. {``TTC Subway Delay Data.''}
\url{https://open.toronto.ca/dataset/ttc-subway-delay-data/}.

\bibitem[\citeproctext]{ref-diaetal}
Dia, Hussein, Michael Taylor, John Stone, Sekhar Somenahalli, and
Stephen Cook. 2019. {``Low Carbon Urban Mobility.''} In, 259--85.
\url{https://doi.org/10.1007/978-981-13-7940-6_14}.

\bibitem[\citeproctext]{ref-janitor}
Firke, Sam. 2023. \emph{Janitor: Simple Tools for Examining and Cleaning
Dirty Data}. \url{https://CRAN.R-project.org/package=janitor}.

\bibitem[\citeproctext]{ref-odt}
Gelfand, Sharla. 2022. \emph{Opendatatoronto: Access the City of Toronto
Open Data Portal}.
\url{https://CRAN.R-project.org/package=opendatatoronto}.

\bibitem[\citeproctext]{ref-lubridate}
Grolemund, Garrett, and Hadley Wickham. 2011. {``Dates and Times Made
Easy with {lubridate}.''} \emph{Journal of Statistical Software} 40 (3):
1--25. \url{https://www.jstatsoft.org/v40/i03/}.

\bibitem[\citeproctext]{ref-MartinWittmanLi}
Martin, Layla, Michael Wittmann, and Xinyu Li. 2021. {``The Influence of
Public Transport Delays on Mobility on Demand Services.''}
\emph{Electronics} 10 (4).
\url{https://www.mdpi.com/2079-9292/10/4/379}.

\bibitem[\citeproctext]{ref-citeR}
R Core Team. 2022. \emph{R: A Language and Environment for Statistical
Computing}. Vienna, Austria: R Foundation for Statistical Computing.
\url{https://www.R-project.org/}.

\bibitem[\citeproctext]{ref-testthat}
Wickham, Hadley. 2011. {``Testthat: Get Started with Testing.''}
\emph{The R Journal} 3: 5--10.
\url{https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf}.

\bibitem[\citeproctext]{ref-ggplot}
---------. 2016. \emph{Ggplot2: Elegant Graphics for Data Analysis}.
Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.

\bibitem[\citeproctext]{ref-tidyverse}
Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy
D'Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019.
{``Welcome to the {tidyverse}.''} \emph{Journal of Open Source Software}
4 (43): 1686. \url{https://doi.org/10.21105/joss.01686}.

\bibitem[\citeproctext]{ref-dplyr}
Wickham, Hadley, Romain François, Lionel Henry, Kirill Müller, and Davis
Vaughan. 2023. \emph{Dplyr: A Grammar of Data Manipulation}.
\url{https://CRAN.R-project.org/package=dplyr}.

\end{CSLReferences}

\end{document}

### 2024-01-24 scripts/01-download_data.R

@@ -7,7 +7,6 @@
 
 #### Workspace setup ####
 library(opendatatoronto)
-library(readxl)
 library(tidyverse)
 
 

### 2024-01-25 outputs/paper/paper.qmd
Cleaned data and started writing paper.

@@
 thanks: "Code and data are available at: https://github.com/lcarnegie/OpenDataToronto.
-Thank you to Nescafé coffee, Rajan Maghera, Egg McMuffins, Hannah Yu, and Sehar Bajwa for your love and support - could not have done it without you.
+Thank you to Nescafé coffee, Rajan Maghera, Hannah Yu, and Sehar Bajwa for your love and support - could not have done it without you.
 No thanks go to Rohan in particular for not pointing out all the R he wanted us to learn that was in his textbook...
@@
 library(kableExtra)
+library(knitr)
 ```
@@
 A key area of improvement in Canadian sustainability is in increasing the use of public transportation in big cities like Toronto.
-This makes sense since cars account for about 80% of transportation emissions [@diaetal].
-However, a major drawback preventing the widespread use of public transportation is the existence of delays within the transit system.
-These delays incentivize citizens toward using less sustainable alternatives like private cars, leading to congestion on highways and consequent increased emissions.
+This makes sense, since cars account for about 80% of transportation emissions [@diaetal].
+A major drawback preventing the widespread use of public transportation is the existence of delays within transit systems.
+These delays push citizens toward using less sustainable alternatives like private cars, leading to congestion on highways and consequent increased emissions.
 The lack of timely transit services have a demonstrated impact on traffic.
@@
 estimated a whopping 47 percent increase in highway delay when transit service ceased completely in a big city such as Los Angeles [@anderson].
-It is clear to see, then, that it is imperative that policy makers understand and fix these delays to improve sustainability.
+Learning from this example, it is imperative that policy makers understand and fix these delays to improve sustainability in Toronto.
 
-The "TTC", or Toronto Transit Commission, has been Toronto's transit system since 1921, serving the city population with subways, streetcars, buses, along with smaller offerings.
-The openly available data surrounding the delays of these various modes of transport offers opportunity to investigate the root causes of the longer-lasting delays.
-Understanding what caused a delay, when it was caused and where it was caused could give the TTC more insight into how to deliver their services to the public more efficiently and effectively, setting an examples of what an effective transit agency could look like.
+The Toronto Transit Commission (TTC) has been Toronto's transit system since 1921, serving the city population with a variety of transportation methods, particularly subways, streetcars, and buses.
+The openly available data surrounding the delays of these various modes of transport offers the opportunity to investigate the root causes of these delays and prompt discussion as how to rectify them.
+Understanding the common causes of a delay, when a delay is typically caused and where it was caused gives the TTC more insight into how to deliver their services to the public more efficiently and effectively, setting an improved example of what an effective transit agency looks like.
 
-This paper aims to find relationships and patterns of public transit delay time across various genres of transportation.
-First, the data source and analysis employed are covered.
-Section Three then critically examines the data, delving deeper to arrive at various conclusions, as well as discussing implications and proposing new areas of exploration.
+This paper aims to find patterns of public transit delay time across various genres of transportation.
+First, the data source and analysis employed are covered in Section Two.
+Section Three then critically examines the data, delving deeper to arrive at various conclusions, as well as discussing implications and proposing new areas of exploration for this data.
 
@@
 Then, the data was cleaned using R [@citeR], tidyverse [@tidyverse] and it's associated packages.
-After combining the datasets together, variables common between datasets and relevant to the analysis were selected.
+After combining the datasets together, variables common between datasets and relevant to the analysis were selected, leading to a dataset with 93,569 observations.
 In the case of this analysis, the variables included were date (year, month, day), day of the week, Location of Delay Incident, Cause of Delay, the time delay (in minutes), Vehicle (bus, subway, streetcar), as well as the route/line (simplified to 'line') the vehicle was on when the incident happened.
-(Perhaps add seasonal analysis??)
-
-## General Delay Analysis {#subsec1-data}
 
@@
 
-sample_table <- head(cleaned_transit_data, 5)
+```
 
-knitr::kable(sample_table)
+```{r}
+#| message: false
+#| echo: false
+#| label: summary_stats_table
+#| tbl-cap: Sample of Cleaned Delay Data
+#| fig-align: center
+
+cleaned_transit_data <- read_csv(here::here("outputs/data/cleaned_dataset.csv"))
+
+sample_table <- cleaned_transit_data |>
+  rename(Vehicle = vehicle) |>
+  rename(`Delay (Minutes)` = `Min Delay`) |>
+  rename(Reason = Incident) |>
+  rename(`Route/Line` = Line) |>
+  arrange(Date)
+
+sample_table <- head(sample_table, 5)
+
+#Scaling taken from: 
+#https://stackoverflow.com/questions/56254631/table-way-too-wide-to-fit-in-markdown-generated-pdf
+kable(sample_table) |> kable_styling(latex_options = c("striped", "scale_down"))
+```
+
+```{r}
+#| message: false
+#| echo: false
+#| label: common-reasons-delay
+#| tbl-cap: Most Common Reasons for Delay on TTC
+
+#Bar chart of Reasons for delay vs. # of incidents
+
+cleaned_transit_data <- read_csv(here::here("outputs/data/cleaned_dataset.csv"))
+
+cleaned_transit_data |>
+  count(Incident) |>
+  ggplot(aes(x = n, y = reorder(Incident, n))) +
+  geom_bar(stat="identity", fill="grey50") +
+  labs(title="Most Common Reasons for Delay", x="Count", y="Incident Type") +
+  theme(axis.text.x = element_text(angle = 30, hjust = 0.5, vjust = 0.5),
+        panel.background = element_blank(),
+        axis.line.y=element_line(color="black"),
+        axis.ticks.y=element_blank(),
+        panel.grid.major.x=element_line(color="grey80"),
+        panel.grid.minor=element_blank())
+
+```
+
+```{r}
+#| message: false
+#| echo: false
+#| label: monthly-delay-times
+#| tbl-cap: Average Monthy Delay Times
+#Line chart of how avg. delay time changes over the months
+
+month_dataset <- cleaned_transit_data |> mutate(month(Date)) |> rename(cleaned_transit_data$month, `month` = `month(Date)`)
+
+# #Calculate avg. delays 
+avg_delay <- month_dataset |> group_by(month, vehicle) |> summarise(`Min Delay` = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")
+
+# #convert month to factor (categorical data)
+avg_delay$month <- factor(avg_delay$month, levels = 1:12, labels = month.abb[1:12])
+
+# #plot the graph
+avg_delay |>
+  ggplot(aes(x=month, y=`Min Delay`, color=vehicle, group=vehicle)) +
+  geom_line() +
+  labs(title="Average Monthly Delays by Vehicle", x="Month", y="Monthly Avg. Delay") +
+  scale_color_manual(values=c("blue","green", "red"), labels=c("Bus","Streetcar", "Subway")) +
+  theme_minimal()
+```
+
+```{r}
+#| message: false
+#| echo: false
+#| label: mode-longest-delays
+#| tbl-cap: Most Common Reasons for Delay on TTC
+#Bar graph of mode of transport with longest delays
+
+avg_time_delayed_vehicle <- cleaned_transit_data |> group_by(vehicle) |> summarise(avg_delay = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")
+
+avg_time_delayed_vehicle |>
+  ggplot(mapping = aes(x = vehicle, y = avg_delay)) +
+  geom_col() + 
+  labs(title="Average Delay by Vehicle", x="Vehicle Type", y="Average Delay (mins)") +
+  theme_minimal()
+```
+
+```{r}
+#| message: false
+#| echo: false
+#| label: most-incidents-vehicle
+#| tbl-cap: Most Incidents by Vehicle
+#Bar graph of mode of transport with most number of delays
+
+num_incidents_vehicle <- cleaned_transit_data |> group_by(vehicle) |> summarise(num_incidents = n(), .groups = "drop")
+
+num_incidents_vehicle |>
+  ggplot(mapping = aes(x = vehicle, y = num_incidents)) +
+  geom_col() + 
+  labs(title="Number of Incidents by Vehicle", x="Vehicle Type", y="Number of Incidents") +
+  theme_minimal()
+
+```
+
+```{r}
+#| message: false
+#| echo: false
+#| label: transport-delay-time
+#| tbl-cap: Most Common Reasons for Delay on TTC
+#Bar graph of mode of transport with most number of delays
+
+# Inspired by: https://stackoverflow.com/questions/34280140/order-bar-graph-in-order-in-this-case-as-per-days-of-week
+
+clean_transit_data <- cleaned_transit_data
+
+# Define the order of days
+days_order <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")
+
+# Convert 'Day' column to a factor with specific levels
+clean_transit_data$Day <- factor(clean_transit_data$Day, levels = days_order)
+
+# Now create the plot
+num_incidents_vehicle <- clean_transit_data |> arrange(Day) |> group_by(Day) |> summarise(num_incidents = n(), .groups = "drop")
+
+num_incidents_vehicle |>
+  ggplot(mapping = aes(x = Day, y = num_incidents)) +
+  geom_col() +
+  labs(title="Count of Delays by Day of Week", x="Day of Week", y="Incident Count") +
+  theme_minimal()
+
+```
+
+```{r}
+#| message: false
+#| echo: false
+#| label: transport-delay-number
+#| tbl-cap: Most Common Reasons for Delay on TTC
+#| 
+# Define the order of days
+days_order <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")
+
+# Convert 'Day' column to a factor with specific levels
+clean_transit_data$Day <- factor(clean_transit_data$Day, levels = days_order)
+
+avg_time_delayed_vehicle <- clean_transit_data |> group_by(Day) |> summarise(avg_delay = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")
+
+avg_time_delayed_vehicle |>
+  ggplot(mapping = aes(x = Day, y = avg_delay)) +
+  geom_col() + 
+  labs(title="Average Delay by Day of Week", x="Vehicle Type", y="Average Delay (mins)") +
+  theme_minimal()
 ```

### 2024-01-25 outputs/paper/paper.tex
Cleaned data and started writing paper.

@@
 https://github.com/lcarnegie/OpenDataToronto. Thank you to Nescafé
-coffee, Rajan Maghera, Egg McMuffins, Hannah Yu, and Sehar Bajwa for
-your love and support - could not have done it without you. No thanks go
-to Rohan in particular for not pointing out all the R he wanted us to
-learn that was in his textbook\ldots{} sigh. I will try the appendix
-cocktails someday, though. :)}}
+coffee, Rajan Maghera, Hannah Yu, and Sehar Bajwa for your love and
+support - could not have done it without you. No thanks go to Rohan in
+particular for not pointing out all the R he wanted us to learn that was
+in his textbook\ldots{} sigh. I will try the appendix cocktails someday,
+though. :)}}
 \usepackage{etoolbox}
@@
 \author{Luca Carnegie}
-\date{January 24, 2024}
+\date{January 25, 2024}
 
@@
 the use of public transportation in big cities like Toronto. This makes
-sense since cars account for about 80\% of transportation emissions (Dia
-et al. 2019). However, a major drawback preventing the widespread use of
-public transportation is the existence of delays within the transit
-system. These delays incentivize citizens toward using less sustainable
-alternatives like private cars, leading to congestion on highways and
-consequent increased emissions. The lack of timely transit services have
-a demonstrated impact on traffic. One study by Anderson et al.~estimated
-a whopping 47 percent increase in highway delay when transit service
-ceased completely in a big city such as Los Angeles (Anderson 2014). It
-is clear to see, then, that it is imperative that policy makers
-understand and fix these delays to improve sustainability.
+sense, since cars account for about 80\% of transportation emissions
+(Dia et al. 2019). A major drawback preventing the widespread use of
+public transportation is the existence of delays within transit systems.
+These delays push citizens toward using less sustainable alternatives
+like private cars, leading to congestion on highways and consequent
+increased emissions. The lack of timely transit services have a
+demonstrated impact on traffic. One study by Anderson et al.~estimated a
+whopping 47 percent increase in highway delay when transit service
+ceased completely in a big city such as Los Angeles (Anderson 2014).
+Learning from this example, it is imperative that policy makers
+understand and fix these delays to improve sustainability in Toronto.
 
-The ``TTC'', or Toronto Transit Commission, has been Toronto's transit
-system since 1921, serving the city population with subways, streetcars,
-buses, along with smaller offerings. The openly available data
-surrounding the delays of these various modes of transport offers
-opportunity to investigate the root causes of the longer-lasting delays.
-Understanding what caused a delay, when it was caused and where it was
-caused could give the TTC more insight into how to deliver their
-services to the public more efficiently and effectively, setting an
-examples of what an effective transit agency could look like.
+The Toronto Transit Commission (TTC) has been Toronto's transit system
+since 1921, serving the city population with a variety of transportation
+methods, particularly subways, streetcars, and buses. The openly
+available data surrounding the delays of these various modes of
+transport offers the opportunity to investigate the root causes of these
+delays and prompt discussion as how to rectify them. Understanding the
+common causes of a delay, when a delay is typically caused and where it
+was caused gives the TTC more insight into how to deliver their services
+to the public more efficiently and effectively, setting an improved
+example of what an effective transit agency looks like.
 
-This paper aims to find relationships and patterns of public transit
-delay time across various genres of transportation. First, the data
-source and analysis employed are covered. Section Three then critically
+This paper aims to find patterns of public transit delay time across
+various genres of transportation. First, the data source and analysis
+employed are covered in Section Two. Section Three then critically
 examines the data, delving deeper to arrive at various conclusions, as
-well as discussing implications and proposing new areas of exploration.
+well as discussing implications and proposing new areas of exploration
+for this data.
 
@@
 associated packages. After combining the datasets together, variables
-common between datasets and relevant to the analysis were selected. In
-the case of this analysis, the variables included were date (year,
-month, day), day of the week, Location of Delay Incident, Cause of
-Delay, the time delay (in minutes), Vehicle (bus, subway, streetcar), as
-well as the route/line (simplified to `line') the vehicle was on when
-the incident happened. (Perhaps add seasonal analysis??)
-
-\subsection{General Delay Analysis}\label{subsec1-data}
+common between datasets and relevant to the analysis were selected,
+leading to a dataset with 93,569 observations. In the case of this
+analysis, the variables included were date (year, month, day), day of
+the week, Location of Delay Incident, Cause of Delay, the time delay (in
+minutes), Vehicle (bus, subway, streetcar), as well as the route/line
+(simplified to `line') the vehicle was on when the incident happened.
 
@@
 
+}
+
+\end{table}%
+
+\begin{table}
+\caption{Sample of Cleaned Delay Data}\tabularnewline
+
+\centering
+\resizebox{\linewidth}{!}{
 \begin{tabular}{l|l|l|l|r|l|l|r}
 \hline
-Date & Time & Day & vehicle & Line & Location & Incident & Min Delay\\
+Date & Time & Day & Vehicle & Route/Line & Location & Reason & Delay (Minutes)\\
 \hline
-2023-01-01 & 02:30:00 & Sunday & Bus & 91 & WOODBINE AND MORTIMER & Diversion & 81\\
+\cellcolor{gray!6}{2023-01-01} & \cellcolor{gray!6}{02:30:00} & \cellcolor{gray!6}{Sunday} & \cellcolor{gray!6}{Bus} & \cellcolor{gray!6}{91} & \cellcolor{gray!6}{WOODBINE AND MORTIMER} & \cellcolor{gray!6}{Diversion} & \cellcolor{gray!6}{81}\\
 \hline
@@
 \hline
-2023-01-01 & 03:06:00 & Sunday & Bus & 35 & JANE STATION & Cleaning & 30\\
+\cellcolor{gray!6}{2023-01-01} & \cellcolor{gray!6}{03:06:00} & \cellcolor{gray!6}{Sunday} & \cellcolor{gray!6}{Bus} & \cellcolor{gray!6}{35} & \cellcolor{gray!6}{JANE STATION} & \cellcolor{gray!6}{Cleaning} & \cellcolor{gray!6}{30}\\
 \hline
@@
 \hline
-2023-01-01 & 03:43:00 & Sunday & Bus & 85 & MEADOWALE LOOP & Security & 1\\
+\cellcolor{gray!6}{2023-01-01} & \cellcolor{gray!6}{03:43:00} & \cellcolor{gray!6}{Sunday} & \cellcolor{gray!6}{Bus} & \cellcolor{gray!6}{85} & \cellcolor{gray!6}{MEADOWALE LOOP} & \cellcolor{gray!6}{Security} & \cellcolor{gray!6}{1}\\
 \hline
-\end{tabular}
+\end{tabular}}
+\end{table}
 
-}
+\includegraphics{paper_files/figure-pdf/common-reasons-delay-1.pdf}
 
-\end{table}%
+\begin{verbatim}
+Warning: Unknown or uninitialised column: `month`.
+\end{verbatim}
+
+\includegraphics{paper_files/figure-pdf/monthly-delay-times-1.pdf}
+
+\includegraphics{paper_files/figure-pdf/mode-longest-delays-1.pdf}
+
+\includegraphics{paper_files/figure-pdf/most-incidents-vehicle-1.pdf}
+
+\includegraphics{paper_files/figure-pdf/transport-delay-time-1.pdf}
+
+\includegraphics{paper_files/figure-pdf/transport-delay-number-1.pdf}
 

### 2024-01-25 scripts/02-data_cleaning.R

@@ -96,7 +96,7 @@ combined_data <- combined_data |> mutate(
       "Late Entering Service" ~ "Late Entering Service", 
       "Overhead" ~ "Overhead", 
       "Rail/Switches" ~ "Rail/Switches", 
-      "NA" ~ "NA"
+      "NA" ~ "N/A"
     )
 )
 

### 2024-01-25 scripts/02-data_cleaning.R
wrote more of the paper

@@ -14,6 +14,7 @@ library(lubridate)
 
 
 #### Clean data ####
+
 raw_data_bus <- read_csv("inputs/data/bus_data.csv")
 raw_data_streetcar <- read_csv("inputs/data/streetcar_data.csv")
 raw_data_subway <- read_csv("inputs/data/subway_data.csv")

### 2024-01-25 outputs/paper/paper.qmd
finished plot code. organized paper.

@@
 abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
-format: pdf
+toc: true
 fig_caption: yes
@@
 bibliography: references.bib
+
+format: pdf
 ---
@@
 library(knitr)
+library(gridExtra)
 ```
@@
 
-This paper aims to find patterns of public transit delay time across various genres of transportation.
+Patterns of public transit delay time are found across various genres of transportation.
 First, the data source and analysis employed are covered in Section Two.
 Section Three then critically examines the data, delving deeper to arrive at various conclusions, as well as discussing implications and proposing new areas of exploration for this data.
+The difference between bus, streetcar, and subway delays is shown to be stark.
+This analysis highlights the need to focus on especially reducing bus delays on several key routes.
 
@@
 
-To investigate transit delays in Toronto, data on bus [@busdata], subway [@subwaydata], and streetcar [@streetcardata] delay incidents for the year 2023 was downloaded using the OpenDataToronto R package [@odt].
-Then, the data was cleaned using R [@citeR], tidyverse [@tidyverse] and it's associated packages.
-After combining the datasets together, variables common between datasets and relevant to the analysis were selected, leading to a dataset with 93,569 observations.
-In the case of this analysis, the variables included were date (year, month, day), day of the week, Location of Delay Incident, Cause of Delay, the time delay (in minutes), Vehicle (bus, subway, streetcar), as well as the route/line (simplified to 'line') the vehicle was on when the incident happened.
-
-```{r}
+```{r data import}
 #| message: false
 #| echo: false
-#| label: tbl-cleaned_data
+#| label: import data
 #| tbl-cap: Sample of Cleaned Delay Data
@@
 
+#make each Line a category, not an integer
+cleaned_transit_data <- cleaned_transit_data |> mutate(Line = factor(Line))
 ```
 
-```{r}
+```{r sample table }
 #| message: false
 #| echo: false
-#| label: summary_stats_table
+#| label: sample_table
 #| tbl-cap: Sample of Cleaned Delay Data
 #| fig-align: center
-
-cleaned_transit_data <- read_csv(here::here("outputs/data/cleaned_dataset.csv"))
 
@@
   rename(`Route/Line` = Line) |>
-  arrange(Date)
-
-sample_table <- head(sample_table, 5)
+  arrange(Date) |>
+  head(5)
 
@@
 #https://stackoverflow.com/questions/56254631/table-way-too-wide-to-fit-in-markdown-generated-pdf
+
 kable(sample_table) |> kable_styling(latex_options = c("striped", "scale_down"))
@@
 
-```{r}
+To investigate transit delays in Toronto, data on bus [@busdata], subway [@subwaydata], and streetcar [@streetcardata] delay incidents for the year 2023 was downloaded using the OpenDataToronto R package [@odt] from the Toronto Open Data Catalogue.
+The data was then cleaned using R [@citeR], tidyverse [@tidyverse] and it's associated packages.
+After combining the datasets together, variables common between datasets and relevant to the analysis were selected, leading to a dataset with 93,569 observations.
+In the case of this analysis, the variables included were date (year, month, day), day of the week, Location of Delay Incident, Cause of Delay, the time delay (in minutes), Vehicle (bus, subway, streetcar), as well as the route/line (simplified to 'line') the vehicle was on when the incident happened.
+
+## Incident Type {#subsec1-data}
+```{r ggplot Common Delay Reasons}
 #| message: false
@@
 
-cleaned_transit_data <- read_csv(here::here("outputs/data/cleaned_dataset.csv"))
+delay_dataset <- cleaned_transit_data
 
-cleaned_transit_data |>
+#Drop N/A incidents (they only apply to subways; subway dataset didn't indicate reasons with entries)
+delay_dataset <- delay_dataset |> filter(Incident != "N/A")
+
+delay_dataset |>
   count(Incident) |>
@@
   geom_bar(stat="identity", fill="grey50") +
-  labs(title="Most Common Reasons for Delay", x="Count", y="Incident Type") +
+  labs(title="Bus and Streetcar Delay Reasons", x="Count", y="Incident Type") +
   theme(axis.text.x = element_text(angle = 30, hjust = 0.5, vjust = 0.5),
@@
         panel.grid.major.x=element_line(color="grey80"),
-        panel.grid.minor=element_blank())
+        panel.grid.minor=element_blank(),
+        plot.title.position = "plot",
+        plot.title = element_text(hjust = 0.5))
+
+```
+Subways didn't have any reason for delay, yet have the most incidence of delay.
+
+## Line/Route {#subsec2-data}
+```{r ggplot top 10 routes with longest avg. delay}
+#| message: false
+#| echo: false
+#| label: lines-longest-delays
+#| tbl-cap: Incidence of Delay, by Route/Line
+
+delay_by_line <- cleaned_transit_data
+
+delay_by_line <- delay_by_line |> 
+                 group_by(Line) |> 
+                 summarise(avg_delay = mean(`Min Delay`, na.rm = TRUE), .groups = "drop") |> 
+                 arrange(desc(avg_delay)) |> 
+                 head(10)
+
+# Plotting
+ggplot(delay_by_line, aes(x = reorder(Line, desc(avg_delay)), y = avg_delay)) +
+  geom_bar(stat = "identity") +
+  labs(x = "Route/Line", y = "Average Delay (Minutes)", title = "Top 10 Routes with Longest Average Delay") +
+  theme_minimal() +
+  theme(plot.title.position = "plot",
+        plot.title = element_text(hjust = 0.5))
+
+```
+```{r ggplot top 10 routes with most delay incidence}
+#| message: false
+#| echo: false
+#| label: lines-most-delays
+#| tbl-cap: Incidence of Delay, by Route/Line
+
+delay_by_line <- cleaned_transit_data
+
+delay_by_line <- delay_by_line |> 
+                 group_by(Line) |> 
+                 summarise(num_delays = n(), .groups = "drop") |> 
+                 arrange(desc(num_delays)) |> 
+                 head(10)
+delay_by_line
+
+# Plotting
+ggplot(delay_by_line, aes(x = reorder(Line, desc(num_delays)), y = num_delays)) +
+  geom_bar(stat = "identity") +
+  labs(x = "Route/Line", y = "Delay Incidence Count", title = "Top 10 Routes with Highest Delay Incidence") +
+  theme_minimal() +
+  theme(plot.title.position = "plot",
+        plot.title = element_text(hjust = 0.5))
 
@@
 
-```{r}
-#| message: false
-#| echo: false
-#| label: monthly-delay-times
-#| tbl-cap: Average Monthy Delay Times
-#Line chart of how avg. delay time changes over the months
-
-month_dataset <- cleaned_transit_data |> mutate(month(Date)) |> rename(cleaned_transit_data$month, `month` = `month(Date)`)
-
-# #Calculate avg. delays 
-avg_delay <- month_dataset |> group_by(month, vehicle) |> summarise(`Min Delay` = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")
-
-# #convert month to factor (categorical data)
-avg_delay$month <- factor(avg_delay$month, levels = 1:12, labels = month.abb[1:12])
-
-# #plot the graph
-avg_delay |>
-  ggplot(aes(x=month, y=`Min Delay`, color=vehicle, group=vehicle)) +
-  geom_line() +
-  labs(title="Average Monthly Delays by Vehicle", x="Month", y="Monthly Avg. Delay") +
-  scale_color_manual(values=c("blue","green", "red"), labels=c("Bus","Streetcar", "Subway")) +
-  theme_minimal()
-```
-
-```{r}
+## Vehicle Type {#subsec3-data}
+```{r ggplot transport with longest avg. delays}
 #| message: false
@@
 ```
-
-```{r}
+```{r ggplot transport with most incidence of delays }
 #| message: false
@@
 #| tbl-cap: Most Incidents by Vehicle
-#Bar graph of mode of transport with most number of delays
+#Bar graph of mode of transport with most incidence of delays
 
@@
 
-```{r}
+## Day of Week {#subsec4-data}
+```{r ggplot Day of Week with Delay incidence}
 #| message: false
@@
 #| tbl-cap: Most Common Reasons for Delay on TTC
-#Bar graph of mode of transport with most number of delays
+#Day of week with most incidence of delay
 
@@
 
-clean_transit_data <- cleaned_transit_data
+dow_data <- cleaned_transit_data
 
@@
 # Convert 'Day' column to a factor with specific levels
-clean_transit_data$Day <- factor(clean_transit_data$Day, levels = days_order)
+dow_data$Day <- factor(dow_data$Day, levels = days_order)
 
 # Now create the plot
-num_incidents_vehicle <- clean_transit_data |> arrange(Day) |> group_by(Day) |> summarise(num_incidents = n(), .groups = "drop")
+num_incidents_vehicle <- dow_data |> arrange(Day) |> group_by(Day) |> summarise(num_incidents = n(), .groups = "drop")
 
@@
   geom_col() +
-  labs(title="Count of Delays by Day of Week", x="Day of Week", y="Incident Count") +
+  labs(title="Incidence of Delay by Day of Week", x="Day of Week", y="Incident Count") +
   theme_minimal()
@@
 ```
-
-```{r}
+```{r ggplot Day of Week with Avg. Delays}
 #| message: false
@@
 #|
+
+dow_data <- cleaned_transit_data
+
 # Define the order of days
@@
 # Convert 'Day' column to a factor with specific levels
-clean_transit_data$Day <- factor(clean_transit_data$Day, levels = days_order)
+dow_data$Day <- factor(dow_data$Day, levels = days_order)
 
-avg_time_delayed_vehicle <- clean_transit_data |> group_by(Day) |> summarise(avg_delay = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")
+#create tibble with avg delay time
+avg_time_delayed_vehicle <- dow_data |> group_by(Day) |> summarise(avg_delay = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")
 
+#plot
 avg_time_delayed_vehicle |>
@@
   geom_col() + 
-  labs(title="Average Delay by Day of Week", x="Vehicle Type", y="Average Delay (mins)") +
+  labs(title="Average Delay Time by Day of Week", x="Vehicle Type", y="Average Delay (mins)") +
+  theme_minimal()
+```
+
+## Time of Year {#subsec5-data}
+```{r ggplot avg delay change monthly}
+#| message: false
+#| echo: false
+#| warning: false
+#| label: monthly-delay-times
+#| tbl-cap: Average Monthy Delay Times
+#Line chart of how avg. delay time changes over the months
+
+# create tibble with month col
+month_dataset <- cleaned_transit_data |> mutate(month(Date)) |> rename(cleaned_transit_data$month, `month` = `month(Date)`)
+
+#Calculate avg. delays 
+avg_delay <- month_dataset |> group_by(month, vehicle) |> summarise(`Min Delay` = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")
+
+# convert month to factor (categorical data)
+avg_delay$month <- factor(avg_delay$month, levels = 1:12, labels = month.abb[1:12])
+
+# plot the graph
+avg_delay |>
+  ggplot(aes(x=month, y=`Min Delay`, color=vehicle, group=vehicle)) +
+  geom_line() +
+  labs(title="Average Monthly Delays by Vehicle", x="Month", y="Monthly Avg. Delay (minutes)") +
+  scale_color_manual(values=c("blue","green", "red"), labels=c("Bus","Streetcar", "Subway")) +
   theme_minimal()
@@
 # Discussion {#sec-discussion}
+
+## Results {#subsec1-discussion}
+
+## Implications {#subsec2-discussion}
+
+## New Areas of Exploration {#subsec3-discussion}
+
+# Conclusion
 
@@
 ---
+
+# Appendix: TTC Route Nomenclature

### 2024-01-26 outputs/paper/paper.qmd
graphs formatted. no writing for data section yet.

@@
 # Data {#sec-data}
-
 ```{r data import}
@@
 
+To investigate transit delays in Toronto, data on bus [@busdata], subway [@subwaydata], and streetcar [@streetcardata] delay incidents for the year 2023 were downloaded using the OpenDataToronto R package [@odt] from the Toronto Open Data Catalogue.
+The data was then cleaned using R [@citeR], tidyverse [@tidyverse] and it's associated packages.
+After combining the datasets together, variables common between datasets and relevant to the analysis were selected, creating a dataset with 93,569 observations.
+For this analysis, the date (year, month, day), time, day of week, vehicle type, route/line, incident location, cause of delay, and time delay (in minutes) were considered.
+
 ```{r sample table }
 #| message: false
+#| warning: false
 #| echo: false
@@
 
-kable(sample_table) |> kable_styling(latex_options = c("striped", "scale_down"))
+kable(sample_table, format = "markdown") |> kable_styling(latex_options = c("scale_down"))
 ```
-
-To investigate transit delays in Toronto, data on bus [@busdata], subway [@subwaydata], and streetcar [@streetcardata] delay incidents for the year 2023 was downloaded using the OpenDataToronto R package [@odt] from the Toronto Open Data Catalogue.
-The data was then cleaned using R [@citeR], tidyverse [@tidyverse] and it's associated packages.
-After combining the datasets together, variables common between datasets and relevant to the analysis were selected, leading to a dataset with 93,569 observations.
-In the case of this analysis, the variables included were date (year, month, day), day of the week, Location of Delay Incident, Cause of Delay, the time delay (in minutes), Vehicle (bus, subway, streetcar), as well as the route/line (simplified to 'line') the vehicle was on when the incident happened.
 
@@
 ```
+
 ```{r ggplot top 10 routes with most delay incidence}
@@
 #| label: lines-most-delays
-#| tbl-cap: Incidence of Delay, by Route/Line
+#| tbl-cap: Most Common Reasons for Delay on TTC
 
+#Routes with highest delay incidence
 delay_by_line <- cleaned_transit_data
@@
                  head(10)
-delay_by_line
+invisible(delay_by_line)
 
@@
 #| label: most-incidents-vehicle
-#| tbl-cap: Most Incidents by Vehicle
+
 #Bar graph of mode of transport with most incidence of delays
 
-num_incidents_vehicle <- cleaned_transit_data |> group_by(vehicle) |> summarise(num_incidents = n(), .groups = "drop")
+most_delayed_vehicle <- cleaned_transit_data
+
+num_incidents_vehicle <- most_delayed_vehicle |> group_by(vehicle) |> summarise(num_incidents = n(), .groups = "drop")
 

### 2024-01-26 outputs/paper/paper.qmd
Finished all the paper writing.

@@
 ---
-title: "Exploring TTC Delays"
-subtitle: "A Multimodal Analysis"
+title: "Transit Delays in the Toronto Transit Commission"
+subtitle: "An Exploratory Analysis with Open Data Toronto"
+
 author: Luca Carnegie
 thanks: "Code and data are available at: https://github.com/lcarnegie/OpenDataToronto.
-Thank you to Nescafé coffee, Rajan Maghera, Hannah Yu, and Sehar Bajwa for your love and support - could not have done it without you.
-No thanks go to Rohan in particular for not pointing out all the R he wanted us to learn that was in his textbook...
-sigh.
-I will try the appendix cocktails someday, though.
-:)"
+Thank you to Nescafé coffee, Rajan Maghera, Hannah Yu, and Sehar Bajwa for your love and support - I could not have done it without you.
+Special thanks go to Rohan Alexander for the immense amount of R he taught me in such a short time and his extremely easy-to-navigate textbook (that has a cocktail appendix, fortunately)."
 date: today
@@
 
-# Data {#sec-data}
+# Data Analysis {#sec-data}
 ```{r data import}
@@
 
-## Incident Type {#subsec1-data}
+## Reason of Delay {#subsec1-data}
 ```{r ggplot Common Delay Reasons}
@@
 ```
-Subways didn't have any reason for delay, yet have the most incidence of delay.
+Subways didn't have any reason for delay, yet have the most incidence of delay. Was dropped from the data.
 
-## Line/Route {#subsec2-data}
+## Day of Week {#subsec2-data}
+```{r ggplot Day of Week with Delay Incidence}
+#| message: false
+#| echo: false
+#| label: transport-delay-time
+#| tbl-cap: Most Common Reasons for Delay on TTC
+#Day of week with most incidence of delay
+
+# Inspired by: https://stackoverflow.com/questions/34280140/order-bar-graph-in-order-in-this-case-as-per-days-of-week
+
+dow_data <- cleaned_transit_data
+
+# Define the order of days
+days_order <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")
+
+# Convert 'Day' column to a factor with specific levels
+dow_data$Day <- factor(dow_data$Day, levels = days_order)
+
+# Now create the plot
+num_incidents_vehicle <- dow_data |> arrange(Day) |> group_by(Day) |> summarise(num_incidents = n(), .groups = "drop")
+
+num_incidents_vehicle |>
+  ggplot(mapping = aes(x = Day, y = num_incidents)) +
+  geom_point() +
+  geom_line(group = 1) +
+  labs(title="Number of Incidents by Day of Week", x="Day of Week", y="Incident Count") +
+  theme_minimal() +
+  ylim(0, max(num_incidents_vehicle$num_incidents) + 1) +
+  theme(plot.title.position = "plot",
+        plot.title = element_text(hjust = 0.5))
+
+```
+```{r ggplot Day of Week with Avg. Delays}
+#| message: false
+#| echo: false
+#| label: transport-delay-number
+#| tbl-cap: Most Common Reasons for Delay on TTC
+#|
+
+dow_data <- cleaned_transit_data
+
+# Define the order of days
+days_order <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")
+
+# Convert 'Day' column to a factor with specific levels
+dow_data$Day <- factor(dow_data$Day, levels = days_order, ordered = TRUE)
+
+#create tibble with avg delay time
+avg_time_delayed_vehicle <- dow_data |> group_by(Day) |> summarise(avg_delay = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")
+
+#plot
+avg_time_delayed_vehicle |>
+  ggplot(mapping = aes(x = Day, y = `avg_delay`)) +
+  geom_point() +
+  geom_line(group = 1) + 
+  labs(title="Average Delay Time by Day of Week", x="Day of Week", y="Average Delay (mins)") +
+  theme_minimal() +
+  ylim(0, max(avg_time_delayed_vehicle$avg_delay) + 1) + 
+  theme(plot.title.position = "plot",
+        plot.title = element_text(hjust = 0.5))
+```
+
+## Vehicle Type {#subsec3-data}
+```{r ggplot transport with longest avg. delays}
+#| message: false
+#| echo: false
+#| label: mode-longest-delays
+#| tbl-cap: Most Common Reasons for Delay on TTC
+#Bar graph of mode of transport with longest delays
+
+avg_time_delayed_vehicle <- cleaned_transit_data |> group_by(vehicle) |> summarise(avg_delay = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")
+
+avg_time_delayed_vehicle |>
+  ggplot(mapping = aes(x = vehicle, y = avg_delay)) +
+  geom_col() + 
+  labs(title="Average Delay by Vehicle", x="Vehicle Type", y="Average Delay (mins)") +
+  theme_minimal() +
+  theme(plot.title.position = "plot",
+      plot.title = element_text(hjust = 0.5))
+```
+```{r ggplot transport with most incidence of delays }
+#| message: false
+#| echo: false
+#| label: most-incidents-vehicle
+
+#Bar graph of mode of transport with most incidence of delays
+
+most_delayed_vehicle <- cleaned_transit_data
+
+num_incidents_vehicle <- most_delayed_vehicle |> group_by(vehicle) |> summarise(num_incidents = n(), .groups = "drop")
+
+num_incidents_vehicle |>
+  ggplot(mapping = aes(x = vehicle, y = num_incidents)) +
+  geom_col() + 
+  labs(title="Number of Incidents by Vehicle", x="Vehicle Type", y="Number of Incidents") +
+  theme_minimal() + 
+   theme(plot.title.position = "plot",
+        plot.title = element_text(hjust = 0.5))
+
+```
+
+## Time of Year {#subsec4-data}
+```{r ggplot incidence change monthly}
+#| message: false
+#| echo: false
+#| warning: false
+#| label: monthly-delay-incidence
+#| tbl-cap: Delay Incidence by Month
+#Line chart of how avg. delay time changes over the months
+
+# create tibble with month col
+month_dataset <- cleaned_transit_data |> mutate(month(Date)) |> rename(cleaned_transit_data$month, `month` = `month(Date)`)
+
+#Calculate avg. delays 
+delay_incidence <- month_dataset |> group_by(month, vehicle) |> summarise(incidence = n(), .groups = "drop")
+
+# convert month to factor (categorical data)
+delay_incidence$month <- factor(delay_incidence$month, levels = 1:12, labels = month.abb[1:12])
+
+# plot the graph
+delay_incidence |>
+  ggplot(aes(x=month, y= `incidence`, color=vehicle, group=vehicle)) +
+  geom_line() +
+  labs(title="Delay Count per Month, by Vehicle", x="Month", y="Count of Delays") +
+  scale_color_manual(values=c("blue","green", "red"), labels=c("Bus","Streetcar", "Subway")) +
+  theme_minimal()
+```
+```{r ggplot avg delay change monthly}
+#| message: false
+#| echo: false
+#| warning: false
+#| label: monthly-delay-times
+#| tbl-cap: Average Monthy Delay Times
+#Line chart of how avg. delay time changes over the months
+
+# create tibble with month col
+month_dataset <- cleaned_transit_data |> mutate(month(Date)) |> rename(cleaned_transit_data$month, `month` = `month(Date)`)
+
+#Calculate avg. delays 
+avg_delay <- month_dataset |> group_by(month, vehicle) |> summarise(`Min Delay` = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")
+
+# convert month to factor (categorical data)
+avg_delay$month <- factor(avg_delay$month, levels = 1:12, labels = month.abb[1:12])
+
+# plot the graph
+avg_delay |>
+  ggplot(aes(x=month, y=`Min Delay`, color=vehicle, group=vehicle)) +
+  geom_line() +
+  labs(title="Monthly Average Delay Time, by Vehicle", x="Month", y="Monthly Avg. Delay (minutes)") +
+  scale_color_manual(values=c("blue","green", "red"), labels=c("Bus","Streetcar", "Subway"))+
+  theme_minimal() +
+  theme(plot.title.position = "plot",
+        plot.title = element_text(hjust = 0.5))
+```
+
+## Line/Route {#subsec5-data}
 ```{r ggplot top 10 routes with longest avg. delay}
@@
 ```
-
 ```{r ggplot top 10 routes with most delay incidence}
@@
 
-## Vehicle Type {#subsec3-data}
-```{r ggplot transport with longest avg. delays}
-#| message: false
-#| echo: false
-#| label: mode-longest-delays
-#| tbl-cap: Most Common Reasons for Delay on TTC
-#Bar graph of mode of transport with longest delays
-
-avg_time_delayed_vehicle <- cleaned_transit_data |> group_by(vehicle) |> summarise(avg_delay = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")
-
-avg_time_delayed_vehicle |>
-  ggplot(mapping = aes(x = vehicle, y = avg_delay)) +
-  geom_col() + 
-  labs(title="Average Delay by Vehicle", x="Vehicle Type", y="Average Delay (mins)") +
-  theme_minimal()
-```
-```{r ggplot transport with most incidence of delays }
-#| message: false
-#| echo: false
-#| label: most-incidents-vehicle
-
-#Bar graph of mode of transport with most incidence of delays
-
-most_delayed_vehicle <- cleaned_transit_data
-
-num_incidents_vehicle <- most_delayed_vehicle |> group_by(vehicle) |> summarise(num_incidents = n(), .groups = "drop")
-
-num_incidents_vehicle |>
-  ggplot(mapping = aes(x = vehicle, y = num_incidents)) +
-  geom_col() + 
-  labs(title="Number of Incidents by Vehicle", x="Vehicle Type", y="Number of Incidents") +
-  theme_minimal()
-
-```
-
-## Day of Week {#subsec4-data}
-```{r ggplot Day of Week with Delay incidence}
-#| message: false
-#| echo: false
-#| label: transport-delay-time
-#| tbl-cap: Most Common Reasons for Delay on TTC
-#Day of week with most incidence of delay
-
-# Inspired by: https://stackoverflow.com/questions/34280140/order-bar-graph-in-order-in-this-case-as-per-days-of-week
-
-dow_data <- cleaned_transit_data
-
-# Define the order of days
-days_order <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")
-
-# Convert 'Day' column to a factor with specific levels
-dow_data$Day <- factor(dow_data$Day, levels = days_order)
-
-# Now create the plot
-num_incidents_vehicle <- dow_data |> arrange(Day) |> group_by(Day) |> summarise(num_incidents = n(), .groups = "drop")
-
-num_incidents_vehicle |>
-  ggplot(mapping = aes(x = Day, y = num_incidents)) +
-  geom_col() +
-  labs(title="Incidence of Delay by Day of Week", x="Day of Week", y="Incident Count") +
-  theme_minimal()
-
-```
-```{r ggplot Day of Week with Avg. Delays}
-#| message: false
-#| echo: false
-#| label: transport-delay-number
-#| tbl-cap: Most Common Reasons for Delay on TTC
-#|
-
-dow_data <- cleaned_transit_data
-
-# Define the order of days
-days_order <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")
-
-# Convert 'Day' column to a factor with specific levels
-dow_data$Day <- factor(dow_data$Day, levels = days_order)
-
-#create tibble with avg delay time
-avg_time_delayed_vehicle <- dow_data |> group_by(Day) |> summarise(avg_delay = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")
-
-#plot
-avg_time_delayed_vehicle |>
-  ggplot(mapping = aes(x = Day, y = avg_delay)) +
-  geom_col() + 
-  labs(title="Average Delay Time by Day of Week", x="Vehicle Type", y="Average Delay (mins)") +
-  theme_minimal()
-```
-
-## Time of Year {#subsec5-data}
-```{r ggplot avg delay change monthly}
-#| message: false
-#| echo: false
-#| warning: false
-#| label: monthly-delay-times
-#| tbl-cap: Average Monthy Delay Times
-#Line chart of how avg. delay time changes over the months
-
-# create tibble with month col
-month_dataset <- cleaned_transit_data |> mutate(month(Date)) |> rename(cleaned_transit_data$month, `month` = `month(Date)`)
-
-#Calculate avg. delays 
-avg_delay <- month_dataset |> group_by(month, vehicle) |> summarise(`Min Delay` = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")
-
-# convert month to factor (categorical data)
-avg_delay$month <- factor(avg_delay$month, levels = 1:12, labels = month.abb[1:12])
-
-# plot the graph
-avg_delay |>
-  ggplot(aes(x=month, y=`Min Delay`, color=vehicle, group=vehicle)) +
-  geom_line() +
-  labs(title="Average Monthly Delays by Vehicle", x="Month", y="Monthly Avg. Delay (minutes)") +
-  scale_color_manual(values=c("blue","green", "red"), labels=c("Bus","Streetcar", "Subway")) +
-  theme_minimal()
-```
-
 # Discussion {#sec-discussion}
@@
 
-# Conclusion
-
+# Appendix 
 # References
@@
 ---
-
-# Appendix: TTC Route Nomenclature

### 2024-01-26 outputs/paper/paper.qmd
Paper basically finished

@@
 ---
-title: "Transit Delays in the Toronto Transit Commission"
+title: "Bring More Buses: Transit Delays in the Toronto Transit Commission"
 subtitle: "An Exploratory Analysis with Open Data Toronto"
-
 author: Luca Carnegie
-thanks: "Code and data are available at: https://github.com/lcarnegie/OpenDataToronto.
+thanks: "Code and data are available at: https://github.com/lcarnegie/TTCTransitAnalysis.
 Thank you to Nescafé coffee, Rajan Maghera, Hannah Yu, and Sehar Bajwa for your love and support - I could not have done it without you.
@@
 This makes sense, since cars account for about 80% of transportation emissions [@diaetal].
-A major drawback preventing the widespread use of public transportation is the existence of delays within transit systems.
+However, a major drawback preventing the widespread use of public transportation is the existence of delays within the transit system.
 These delays push citizens toward using less sustainable alternatives like private cars, leading to congestion on highways and consequent increased emissions.
@@
 One study by Anderson et al.
-estimated a whopping 47 percent increase in highway delay when transit service ceased completely in a big city such as Los Angeles [@anderson].
-Learning from this example, it is imperative that policy makers understand and fix these delays to improve sustainability in Toronto.
+estimated a whopping 47 percent increase in highway delay when transit service ceases completely in Los Angeles [@anderson], a large city like Toronto.
+It is then imperative that policy makers understand and fix these delays to improve sustainability.
 
-The Toronto Transit Commission (TTC) has been Toronto's transit system since 1921, serving the city population with a variety of transportation methods, particularly subways, streetcars, and buses.
-The openly available data surrounding the delays of these various modes of transport offers the opportunity to investigate the root causes of these delays and prompt discussion as how to rectify them.
-Understanding the common causes of a delay, when a delay is typically caused and where it was caused gives the TTC more insight into how to deliver their services to the public more efficiently and effectively, setting an improved example of what an effective transit agency looks like.
+The "TTC", or Toronto Transit Commission, has been Toronto's transit system since 1921, serving the city population with mainly subways, streetcars, buses, along with smaller-scale transport offerings.
+The openly available data surrounding the delays of these various modes of transport offers opportunity to investigate the root causes of the longer-lasting delays.
+In addition, delays give an idea of  Understanding what caused a delay, when it was caused and where it was caused could give the TTC more insight into how to deliver their services to the public more efficiently and effectively, setting an examples of what an effective transit agency could look like.
 
-Patterns of public transit delay time are found across various genres of transportation.
-First, the data source and analysis employed are covered in Section Two.
-Section Three then critically examines the data, delving deeper to arrive at various conclusions, as well as discussing implications and proposing new areas of exploration for this data.
+This paper finds relationships and patterns of public transit delay time across various genres of transportations.
+The paper is structured in that Section Two covers the data source, and analysis employed.
+Section Three critically examines the data, comes to several conclusions and proposes new areas of exploration of this dataset.
 The difference between bus, streetcar, and subway delays is shown to be stark.
-This analysis highlights the need to focus on especially reducing bus delays on several key routes.
+This analysis highlights the need to focus on especially on reducing bus delays on several key routes to improve efficiency and effectiveness of service.
 
 # Data Analysis {#sec-data}
-```{r data import}
+```{r data-import}
 #| message: false
 #| echo: false
-#| label: import data
-#| tbl-cap: Sample of Cleaned Delay Data
-#| fig-align: center
 
@@
 
-To investigate transit delays in Toronto, data on bus [@busdata], subway [@subwaydata], and streetcar [@streetcardata] delay incidents for the year 2023 were downloaded using the OpenDataToronto R package [@odt] from the Toronto Open Data Catalogue.
-The data was then cleaned using R [@citeR], tidyverse [@tidyverse] and it's associated packages.
-After combining the datasets together, variables common between datasets and relevant to the analysis were selected, creating a dataset with 93,569 observations.
-For this analysis, the date (year, month, day), time, day of week, vehicle type, route/line, incident location, cause of delay, and time delay (in minutes) were considered.
+To investigate transit delays in the TTC, recorded data on bus [@busdata] , streetcar [@streetcardata], and subway [@subwaydata] delay incidents for the year 2023 were obtained using the OpenDataToronto R package [@odt], providing data from the Toronto Open Data Catalogue.
+All the datasets were current as of January 18th, 2024.
 
-```{r sample table }
+All the data was then cleaned using R [@citeR] and it's tidyverse [@tidyverse], janitor [@janitor], and lubridate [@lubridate] packages.
+Variables common between datasets and relevant to this analysis were selected, creating a combined dataset with 93,569 observations made over the year.
+For this analysis, the date (year, month, day), time, day of week, vehicle type, route/line, incident location, cause of delay, and time delay gap (in minutes) were made of interest.
+A sample of the cleaned dataset is shown in [@tbl-sample].
+In the analysis itself, the knitr [@knitr] and ggplot2 [@ggplot] R packages were used to create the tables and graphs.
+
+```{r tbl-sample}
 #| message: false
+#| echo: false
 #| warning: false
-#| echo: false
-#| label: sample_table
-#| tbl-cap: Sample of Cleaned Delay Data
+#| tbl-cap: Delay Incidence by Month
 #| fig-align: center
@@
 
+The dataset, primarily categorical, was analyzed on a variable-by-variable basis.
+Each variable, barring ‘delay reason’, was examined in two ways: through its relationship with delay occurrences and its impact on the average delay duration, measured in minutes.
+
 ## Reason of Delay {#subsec1-data}
-```{r ggplot Common Delay Reasons}
+
+First, we assess the various recorded reasons of a particular delay on the TTC and their impacts.
+In @fig-reasons-delay, we observe several key groupings of delay reasons for buses and streetcars.
+Before analyzing further, however, it is of important note that the reasons for subway delay were not recorded during data collection, subways must be omitted from this particular chart.
+
+```{r fig-reasons-delay}
 #| message: false
 #| echo: false
-#| label: common-reasons-delay
-#| tbl-cap: Most Common Reasons for Delay on TTC
+#| warning: false
+#| tbl-cap: Distribution of Bus and Streetcar Delay Reasons 
+#| fig-align: center
 
@@
 delay_dataset <- delay_dataset |> filter(Incident != "N/A")
+
+summarized <- delay_dataset |> group_by(Incident) |> summarise(count = n(), .groups = "drop")
 
@@
 ```
-Subways didn't have any reason for delay, yet have the most incidence of delay. Was dropped from the data.
+
+In @fig-reasons-delay, Mechanical Issues, Operational Issues, and Route Diversion emerge as the top three causes of delays in streetcars and buses across the TTC.
+The incidence of these delays varies significantly.
+In particular, there 6682 more mechanical delays than operational delays and 13083 more mechanical delays than diversion delays.
 
 ## Day of Week {#subsec2-data}
-```{r ggplot Day of Week with Delay Incidence}
+
+Next, the variation of delay incidents by day of the week was analyzed. Figure @fig-dow-1 shows the number of delay incidents for each day.
+
+```{r fig-dow-1}
 #| message: false
 #| echo: false
-#| label: transport-delay-time
-#| tbl-cap: Most Common Reasons for Delay on TTC
+#| warning: false
+#| tbl-cap: Number of Delay Incidents, aggregated by Day of Week
+#| fig-align: center
 #Day of week with most incidence of delay
@@
 ```
-```{r ggplot Day of Week with Avg. Delays}
+
+The frequency of delays climbs steadily from Sunday to Friday before peaking on Saturday.
+However, there does not appear to be a clear upward or downward trend in delays over the course of the week.
+Further analysis on a weekly basis would be needed to determine if there is a statistically significant pattern in delays based on the day.
+However, the data indicates overall that weekends, particularly Saturdays, experience more delay incidents than weekdays.
+
+```{r fig-dow-2}
 #| message: false
 #| echo: false
-#| label: transport-delay-number
-#| tbl-cap: Most Common Reasons for Delay on TTC
-#|
+#| warning: false
+#| tbl-cap: Average Delay Time, aggregated by Day of Week
+#| fig-align: center
 
@@
 
+Next, we looked at average delay duration by day.
+Figure @fig-dow-2 displays the average delay time in minutes for each day.
+The average remains relatively consistent across the week, hovering around 15 minutes per delay incident.
+
 ## Vehicle Type {#subsec3-data}
-```{r ggplot transport with longest avg. delays}
+
+Next, we graph the data to check for noticeable differences across different genres of transportation. We see a noticeable difference in delays by vehicle in Figures @fig-vehicle-1 and @fig-vehicle-2
+
+```{r fig-vehicle-1}
 #| message: false
 #| echo: false
-#| label: mode-longest-delays
-#| tbl-cap: Most Common Reasons for Delay on TTC
-#Bar graph of mode of transport with longest delays
-
-avg_time_delayed_vehicle <- cleaned_transit_data |> group_by(vehicle) |> summarise(avg_delay = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")
-
-avg_time_delayed_vehicle |>
-  ggplot(mapping = aes(x = vehicle, y = avg_delay)) +
-  geom_col() + 
-  labs(title="Average Delay by Vehicle", x="Vehicle Type", y="Average Delay (mins)") +
-  theme_minimal() +
-  theme(plot.title.position = "plot",
-      plot.title = element_text(hjust = 0.5))
-```
-```{r ggplot transport with most incidence of delays }
-#| message: false
-#| echo: false
-#| label: most-incidents-vehicle
+#| warning: false
+#| tbl-cap: Number of Incidents for each mode of transportation
+#| fig-align: center
 
@@
 
-## Time of Year {#subsec4-data}
-```{r ggplot incidence change monthly}
+Figure @fig-vehicle-1 shows that in 2023, buses had 56207 recorded delay incidents in 2023. This is more than twice as many as subways and close to three times more than streetcars.
+
+```{r fig-vehicle-2}
 #| message: false
@@
 #| warning: false
-#| label: monthly-delay-incidence
+#| tbl-cap: Average Delay Time for each Mode of Transportation
+#| fig-align: center
+
+#Bar graph of mode of transport with longest delays
+
+avg_time_delayed_vehicle <- cleaned_transit_data |> group_by(vehicle) |> summarise(avg_delay = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")
+
+avg_time_delayed_vehicle |>
+  ggplot(mapping = aes(x = vehicle, y = avg_delay)) +
+  geom_col() + 
+  labs(title="Average Delay by Vehicle", x="Vehicle Type", y="Average Delay (mins)") +
+  theme_minimal() +
+  theme(plot.title.position = "plot",
+      plot.title = element_text(hjust = 0.5))
+```
+
+In Figure @fig-vehicle-2, however, we do not see as large a difference in average delay times across different genres of transportation, however.
+Buses had the longest average delay times, at around 20 minutes, but subway had only 3 minutes of delay time on average, making for a fairly sizeable gap delay times.
+Interestingly, although there was less incidence of streetcar delay than subway delay, streetcar routes tended to be delayed longer on average throughout 2023.
+
+## Time of Year {#subsec4-data}
+
+To investigate how TTC delays change throughout the year, Figures @fig-monthly-1 and @fig-monthly-2 were constructed.
+The graphs show monthly changes in delay count and average delay time over 2023.
+Buses, streetcars, and subways are represented by blue, green, and red lines, respectively.
+
+```{r fig-monthly-1}
+#| message: false
+#| echo: false
+#| warning: false
 #| tbl-cap: Delay Incidence by Month
+#| fig-align: center
 #Line chart of how avg. delay time changes over the months
@@
 ```
-```{r ggplot avg delay change monthly}
+
+In @fig-monthly-1 we see how the count of delays changes throughout 2023 for each mode of transportation.
+Buses have the highest incidence of delays throughout year with the count generally trending upward as the year progresses, peaking in November with 5322 incidents.
+Incidence of subway and street car delays are significantly lower, but also trend upwards as the year progresses.
+Overall, this figure presents similar findings to Figure @fig-vehicle-1 in section 2.3 - buses tend to have more incidences of delay than both streetcars and subways overall.
+
+```{r fig-monthly-2}
 #| message: false
@@
 #| warning: false
-#| label: monthly-delay-times
-#| tbl-cap: Average Monthy Delay Times
+#| tbl-cap: Average Delay Time, by Month
 #Line chart of how avg. delay time changes over the months
@@
 
+When looking at changes in average delay times throughout  2023, there are several observations we can make.
+With buses, we observe a slight downward trend in delay time as the year progresses, with delays going from an monthly average of 23.6 minutes to 18.2 minutes by the end of the year.
+Subway delay times stay more or less constant, on average, at around 3.2 minutes, the lowest average delay time of all three.
+With streetcars, however, although delay times tend to trend downward throughout the year, there is a noticeable change in delay times in the summer, with a spike in delay time in the spring and summer months that eclipses even buses in the month of June.
+
 ## Line/Route {#subsec5-data}
-```{r ggplot top 10 routes with longest avg. delay}
+
+Finally, the delays on particular subway lines and bus/streetcar routes were examined.
+Given the large number of different transit routes offered by the TTC, this analysis focuses on the 10 routes with the highest average delay times and highest incidence of delay.
+These findings are illustrated in Figure @fig-highest-1 and Figure @fig-highest-2.
+A short guide to the route number nomenclature of the TTC is provided in the Appendix for the reader's convenience.
+
+```{r fig-highest-1}
 #| message: false
 #| echo: false
-#| label: lines-longest-delays
-#| tbl-cap: Incidence of Delay, by Route/Line
-
-delay_by_line <- cleaned_transit_data
-
-delay_by_line <- delay_by_line |> 
-                 group_by(Line) |> 
-                 summarise(avg_delay = mean(`Min Delay`, na.rm = TRUE), .groups = "drop") |> 
-                 arrange(desc(avg_delay)) |> 
-                 head(10)
-
-# Plotting
-ggplot(delay_by_line, aes(x = reorder(Line, desc(avg_delay)), y = avg_delay)) +
-  geom_bar(stat = "identity") +
-  labs(x = "Route/Line", y = "Average Delay (Minutes)", title = "Top 10 Routes with Longest Average Delay") +
-  theme_minimal() +
-  theme(plot.title.position = "plot",
-        plot.title = element_text(hjust = 0.5))
-
-```
-```{r ggplot top 10 routes with most delay incidence}
-#| message: false
-#| echo: false
-#| label: lines-most-delays
-#| tbl-cap: Most Common Reasons for Delay on TTC
+#| warning: false
+#| tbl-cap: Routes/Lines with Highest Delay Incidence
+#| fig-align: center
 
@@
 
+We observe a mix of bus, streetcar, and subway routes accounting for the highest incidences of delay.
+Interestingly, subway Lines 1 (Yonge-University) and 2 (Bloor-Danforth) occupy the top two positions in terms of delay frequency, with 11628 and 9511 recorded incidents, respectively.
+Following closely are Streetcar Lines 501 (Queen St.) and 504 (King St.), ranking third and fourth.
+Rounding out the top five is Bus Route 32 (Eglinton West), which holds the fifth position in delay incidence, with 1920 recorded incidents.
+
+```{r fig-highest-2}
+#| message: false
+#| echo: false
+#| warning: false
+#| tbl-cap: Routes/Lines with Highest Average Delay Time
+#| fig-align: center
+
+delay_by_line <- cleaned_transit_data
+
+delay_by_line <- delay_by_line |> 
+                 group_by(Line) |> 
+                 summarise(avg_delay = mean(`Min Delay`, na.rm = TRUE), .groups = "drop") |> 
+                 arrange(desc(avg_delay)) |> 
+                 head(10)
+
+# Plotting
+ggplot(delay_by_line, aes(x = reorder(Line, desc(avg_delay)), y = avg_delay)) +
+  geom_bar(stat = "identity") +
+  labs(x = "Route/Line", y = "Average Delay (Minutes)", title = "Top 10 Routes with Longest Average Delay") +
+  theme_minimal() +
+  theme(plot.title.position = "plot",
+        plot.title = element_text(hjust = 0.5))
+
+```
+
+However, the picture changes entirely when routes are looked at in terms of average delay time in @fig-highest-2.
+Here, we see that bus routes entirely make up the 10 routes with the longest average wait times, ranging from 53.4 minutes for Route 162 (Lawrence-Donway) and  up to 133.0 minutes for Route 55 (Warren Park).
+These routes were checked for accuracy in the TTC's route finder [@routes]).
+
 # Discussion {#sec-discussion}
 
-## Results {#subsec1-discussion}
+## Results and Implications {#subsec1-discussion}
 
-## Implications {#subsec2-discussion}
+The analysis of Toronto Transit Commission delays reveals several key findings and implications that could be useful in the public agency’s operations.
+The analysis found that buses experience the longest delay times among all modes of transport, suggesting a dire need to reduce bus waiting times across the board, but particularly on the routes analyzed in section 2.5.
+This could be accomplished in several ways, such as deploying more buses on these routes, or using tactics normally reserved for rail service like bus bridging, where buses are dispatched from scheduled services to act as temporary shuttles [@itanietal].
+However, it is important to consider the relative passenger volume of these routes relative to the system as a whole.
+Perhaps there is such a high delay that the benefit of employing these tactics against the cost to other services provided by the TTC is simply not worth it to the agency.
 
-## New Areas of Exploration {#subsec3-discussion}
+Interestingly, delay incidents tend to occur more frequently on weekends, indicating a need for better trained and/or more response staff on call during those time to address these issues.
+Given the sensitive nature of some incidents, enhanced personnel training could potentially prevent certain situations.
+
+Subways, despite being a popular mode of transport, have the highest incidence of delay.
+This could point to underlying issues in subway service that warrant further investigation.
+As said in Section 2, more data is need to investigate this further, and a good first would be to record the reason for a particular delay when a delay is recorded in the dataset.
+
+The most common reasons for delay are mechanical issues, operations, and route diversions.
+This suggests that subway trains might need to be replaced, and staff could benefit from improved training to reduce delays and flexibility when dealing with route diversions and traffic.
+
+In Toronto's warmer months, April-June particularly, streetcar delay times tend to spike significantly , possibly due to increased pedestrian and cyclist traffic in the downtown core, where streetcars are predominantly located.
+Although delays during that time are unavoidable, it could be worth exploring if training programmes and/or other strategies are being effectively deployed so that streetcars are running efficiently.
+
+## Further Areas of Exploration {#subsec2-discussion}
+
+Future analyses could be more comprehensive with better data on the reasons for subway delays.
+Additionally, given more time and resources to explore additional tools and R packages, location could be another factor of analysis within the scope of TTC delays.
+This would allow us to identify which neighbourhoods experience the most delays, thereby helping to prioritize maintenance crew dispatch and policy decision with respect to adding more vehicles and service to a particular area.
+
+Using data-driven methods to analyze the the TTC's performance of can enable the agency to make decisions that target key routes that need the most assistance, allowing public funds to be used in the most efficient way possible.
+This approach not only improves the efficiency of the TTC but will enhance the commuting experience for the public that depends on it every day to conduct their business in the ci
 
 # Appendix 
+Route nomenclature has gone through many changes throughout the history of the TTC.
+As James Bow, a TTC historian, said to BlogTO, "[The current system] is a kludge of several numbering schemes that were added to or replaced elements of the original scheme, with many elements grandfathered in".
+Hence, he says "there's very little rhyme or reason" to the transit numbering, making it confusing to almost everyone.
+[@flack]
+
+Luckily, in the same interview, conducted in late 2023, he provides a general overview of TTC nomenclature that can provide helpful context when understanding some of the graphs in this paper.
+The numbering is as follows:
+
+- 1-4 - Subway and LRT routes
+- 5-139 - Regular bus routes, assigned willy-nilly
+- 140-159 - Premium express bus services
+- 160-189 - Regular bus routes, assigned willy-nilly
+- 300-399 - Night Bus and Streetcar services
+- 400-499 - Accessible Community Bus services
+- 500-599 - General Streetcar routes.
+- 900-999 - Express bus routes
+
 # References

### 2024-01-26 outputs/paper/paper.qmd
added reference.

@@
 The openly available data surrounding the delays of these various modes of transport offers opportunity to investigate the root causes of the longer-lasting delays.
-In addition, delays give an idea of  Understanding what caused a delay, when it was caused and where it was caused could give the TTC more insight into how to deliver their services to the public more efficiently and effectively, setting an examples of what an effective transit agency could look like.
+Delays are likely the best way to assess a transit agency's performance from both the rider and management perspective, since riders' experience is typically determined through how low waiting and travel times are, while performance of staff is assessed "in terms of
+on-time performance standards (or adherence to schedules)" [@diab2015bus].
+Both of these metrics are highly correlated with delay time.
+Therefore, understanding what caused a delay, when it was caused and where it was caused could give the TTC more insight into how to deliver their services to the public more efficiently and effectively.
 
-This paper finds relationships and patterns of public transit delay time across various genres of transportations.
+This paper finds relationships and patterns of public transit delay time across various genres of transportation.
 The paper is structured in that Section Two covers the data source, and analysis employed.
@@
 # References
-
----
-nocite: |
-  @citeR, @ggplot, @odt, @janitor, @lubridate, @tidyverse, @dplyr, @ggplot, @testthat, @MartinWittmanLi, @diaetal, @anderson,
----

### 2024-01-26 outputs/paper/paper.qmd
paper finished!

@@
 
-The "TTC", or Toronto Transit Commission, has been Toronto's transit system since 1921, serving the city population with mainly subways, streetcars, buses, along with smaller-scale transport offerings.
+The TTC, or Toronto Transit Commission, has been Toronto's transit system since 1921, serving the city population with mainly subways, streetcars, buses, along with smaller-scale transport offerings.
 The openly available data surrounding the delays of these various modes of transport offers opportunity to investigate the root causes of the longer-lasting delays.
@@
 #| warning: false
-#| tbl-cap: Distribution of Bus and Streetcar Delay Reasons 
+#| fig-cap: Distribution of Bus and Streetcar Delay Reasons 
 #| fig-align: center
@@
 
-Next, the variation of delay incidents by day of the week was analyzed. Figure @fig-dow-1 shows the number of delay incidents for each day.
+Next, the variation of delay incidents by day of the week was analyzed. @fig-dow-1-1 shows the number of delay incidents for each day.
 
@@
 #| warning: false
-#| tbl-cap: Number of Delay Incidents, aggregated by Day of Week
+#| fig-cap: Delays, aggregated by Day of Week
+#| fig-subcap: ["Incidences","Average Delay Time (mins)"]
+#| layout-ncol: 2
 #| fig-align: center
@@
 
-```
-
-The frequency of delays climbs steadily from Sunday to Friday before peaking on Saturday.
-However, there does not appear to be a clear upward or downward trend in delays over the course of the week.
-Further analysis on a weekly basis would be needed to determine if there is a statistically significant pattern in delays based on the day.
-However, the data indicates overall that weekends, particularly Saturdays, experience more delay incidents than weekdays.
-
-```{r fig-dow-2}
-#| message: false
-#| echo: false
-#| warning: false
-#| tbl-cap: Average Delay Time, aggregated by Day of Week
-#| fig-align: center
+#Average Time Delay
 
@@
         plot.title = element_text(hjust = 0.5))
+
 ```
 
+The frequency of delays climbs steadily from Sunday to Friday before peaking on Saturday.
+However, there does not appear to be a clear upward or downward trend in delays over the course of the week.
+Further analysis on a weekly basis would be needed to determine if there is a statistically significant pattern in delays based on the day.
+However, the data indicates overall that weekends, particularly Saturdays, experience more delay incidents than weekdays.
+
 Next, we looked at average delay duration by day.
-Figure @fig-dow-2 displays the average delay time in minutes for each day.
+@fig-dow-1-2 displays the average delay time in minutes for each day.
 The average remains relatively consistent across the week, hovering around 15 minutes per delay incident.
@@
 
-Next, we graph the data to check for noticeable differences across different genres of transportation. We see a noticeable difference in delays by vehicle in Figures @fig-vehicle-1 and @fig-vehicle-2
+Next, we graph the data to check for noticeable differences across different genres of transportation. We see a noticeable difference in delays by vehicle in @fig-vehicle-1
 
@@
 #| warning: false
-#| tbl-cap: Number of Incidents for each mode of transportation
+#| fig-cap: Delays for each mode of transportation
 #| fig-align: center
+#| layout-ncol: 2
+#| fig-subcap: ["Number of Incidences","Average Delay Time"]
 
@@
 num_incidents_vehicle <- most_delayed_vehicle |> group_by(vehicle) |> summarise(num_incidents = n(), .groups = "drop")
+
+# kable(num_incidents_vehicle, format = "markdown") |> kable_styling(latex_options = c("scale_down"))
 
@@
 
-```
-
-Figure @fig-vehicle-1 shows that in 2023, buses had 56207 recorded delay incidents in 2023. This is more than twice as many as subways and close to three times more than streetcars.
-
-```{r fig-vehicle-2}
-#| message: false
-#| echo: false
-#| warning: false
-#| tbl-cap: Average Delay Time for each Mode of Transportation
-#| fig-align: center
-
 #Bar graph of mode of transport with longest delays
 
-avg_time_delayed_vehicle <- cleaned_transit_data |> group_by(vehicle) |> summarise(avg_delay = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")
+avg_time_delayed_vehicle <- cleaned_transit_data |> 
+                            group_by(vehicle) |> 
+                            summarise(avg_delay = round(mean(`Min Delay`, na.rm = TRUE), digits = 1), .groups = "drop")
+
+# kable(avg_time_delayed_vehicle, format = "markdown") |> kable_styling(latex_options = c("scale_down"))
 
@@
       plot.title = element_text(hjust = 0.5))
+
 ```
 
-In Figure @fig-vehicle-2, however, we do not see as large a difference in average delay times across different genres of transportation, however.
+@fig-vehicle-1-1 shows that in 2023, buses had 56207 recorded delay incidents in 2023. This is more than twice as many as subways and close to three times more than streetcars.
+
+In @fig-vehicle-1-2, however, we do not see as large a difference in average delay times across different genres of transportation, however.
 Buses had the longest average delay times, at around 20 minutes, but subway had only 3 minutes of delay time on average, making for a fairly sizeable gap delay times.
@@
 
-To investigate how TTC delays change throughout the year, Figures @fig-monthly-1 and @fig-monthly-2 were constructed.
+To investigate how TTC delays change throughout the year, @fig-monthly-1 and @fig-monthly-2 were constructed.
 The graphs show monthly changes in delay count and average delay time over 2023.
@@
 #| warning: false
-#| tbl-cap: Delay Incidence by Month
+#| fig-cap: Delay Incidence by Month
 #| fig-align: center
@@
 Incidence of subway and street car delays are significantly lower, but also trend upwards as the year progresses.
-Overall, this figure presents similar findings to Figure @fig-vehicle-1 in section 2.3 - buses tend to have more incidences of delay than both streetcars and subways overall.
+Overall, this figure presents similar findings to @fig-vehicle-1 in section 2.3 - buses tend to have more incidences of delay than both streetcars and subways overall.
 
@@
 #| warning: false
-#| tbl-cap: Average Delay Time, by Month
+#| fig-cap: Average Delay Time, by Month
 #Line chart of how avg. delay time changes over the months
@@
 Given the large number of different transit routes offered by the TTC, this analysis focuses on the 10 routes with the highest average delay times and highest incidence of delay.
-These findings are illustrated in Figure @fig-highest-1 and Figure @fig-highest-2.
+These findings are illustrated in @fig-highest-1 and @fig-highest-2.
 A short guide to the route number nomenclature of the TTC is provided in the Appendix for the reader's convenience.
@@
 #| warning: false
-#| tbl-cap: Routes/Lines with Highest Delay Incidence
+#| fig-cap: Routes/Lines with Highest Delay Incidence
 #| fig-align: center
@@
 #| warning: false
-#| tbl-cap: Routes/Lines with Highest Average Delay Time
+#| fig-cap: Routes/Lines with Highest Average Delay Time
 #| fig-align: center

### 2024-01-26 outputs/paper/paper.qmd
added claude-2 transcripts

@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "This paper analyzes TTC delay data to understand patterns and causes of delays.
+It finds buses experience the most delays, especially on certain routes, while subways have frequent undocumented delays.
+Mechanical issues, operations, and diversions are most common delay causes.
+The analysis indicates the need to reduce bus delays through more vehicles or tactics like bus bridging on certain routes."
 toc: true
@@
 bibliography: references.bib
-
+title-block-style: none
 format: pdf
@@
 #| fig-subcap: ["Incidences","Average Delay Time (mins)"]
-#| layout-ncol: 2
 #| fig-align: center
@@
 
-Subways, despite being a popular mode of transport, have the highest incidence of delay.
+Subways, being a popular mode of transport, have the highest incidence of delay.
 This could point to underlying issues in subway service that warrant further investigation.
-As said in Section 2, more data is need to investigate this further, and a good first would be to record the reason for a particular delay when a delay is recorded in the dataset.
+As said in Section 2, more data is need to investigate this further, and a good first step to do this would be to record the reason for a particular delay when a delay is recorded in the dataset.
 
 The most common reasons for delay are mechanical issues, operations, and route diversions.
-This suggests that subway trains might need to be replaced, and staff could benefit from improved training to reduce delays and flexibility when dealing with route diversions and traffic.
+This suggests that subway trains might need to be replaced.
+Additionally, staff could benefit from improved training to reduce delays and flexibility when dealing with route diversions and traffic.
 
 In Toronto's warmer months, April-June particularly, streetcar delay times tend to spike significantly , possibly due to increased pedestrian and cyclist traffic in the downtown core, where streetcars are predominantly located.
-Although delays during that time are unavoidable, it could be worth exploring if training programmes and/or other strategies are being effectively deployed so that streetcars are running efficiently.
+Although delays during that time are unavoidable, it could be worth exploring if training programmes and/or other strategies are being effectively deployed so that streetcars are running efficiently while respecting the safety of cyclist in the downtown core.
 
@@
 
-Using data-driven methods to analyze the the TTC's performance of can enable the agency to make decisions that target key routes that need the most assistance, allowing public funds to be used in the most efficient way possible.
-This approach not only improves the efficiency of the TTC but will enhance the commuting experience for the public that depends on it every day to conduct their business in the ci
+Using data-driven methods to analyze the the TTC's performance can enable the agency to make decisions that target key routes that need the most assistance, allowing public funds to be used in the most efficient way possible.
+This approach not only improves the efficiency of the TTC but will enhance the  experience for the commuting public that depends on it every day to conduct their business in the city.
 

### 2024-02-01 outputs/paper/paper.qmd
fixed appendix and refs formatting based on rohan's comments.

@@
 These findings are illustrated in @fig-highest-1 and @fig-highest-2.
-A short guide to the route number nomenclature of the TTC is provided in the Appendix for the reader's convenience.
+A short guide to the route number nomenclature of the TTC is provided in the Appendix (@sec-appendix) for the reader's convenience.
 
@@
 
-# Appendix 
+\newpage
+
+# References
+
+::: {#refs}
+:::
+
+\newpage
+
+# Appendix {#sec-appendix}
+
 Route nomenclature has gone through many changes throughout the history of the TTC.
@@
 - 900-999 - Express bus routes
-
-# References
